Naturalness and Long-Lived Beyond the Standard Model Particles

### 2017-08-03 tex/draft.tex
First real full draft

    SM parameters, such as masses and couplings, are corrected by interactions with virtual particles through loop diagrams. Fig.~\ref{loop_diagrams}, for example, shows interactions that correct the Higgs and electron masses. Because the Higgs is a fundamental scalar, its corrections depend quadratically on the cutoff scale, $\Lambda$, which is the largest energy scale for which the SM is valid. The fermions and gauge bosons, on the other hand, only depend logarithmically\fxnote{check. be able to work out example} on $\Lambda$~\cite{dine_naturalness}. The next section explores the concept of naturalness, a major motivating principle for beyond the standard model (BSM) physics that is directly tied to the relationship between the bare Higgs mass and the terms that correct it.

@@

    All massive particles correct the Higgs mass with terms proportional to $\Lambda^2$~\cite{dine_naturalness}\fxnote{show diagrams and terms?}. Experimentally, the Higgs mass is \SI{125}{\giga\electronvolt}~\cite{cms_higgs, atlas_higgs}, a value that must be achieved by the sum of the bare mass and all corrections. Although corrections differ in sign, the precise cancellation of many extremely large terms to produce a value as small as the Higgs mass is extremely unlikely. In fact, the fine-tuning necessary to produce the Higgs mass is on the order of one part in \num{e30} if the SM is valid up to the Plank scale ($\mPlanck\approx$ \SI{e16}{\giga\electronvolt}), where quantum gravity becomes important~\cite{giudice_naturally}. While it's possible that the bare mass and corrections just so happen to almost perfectly cancel, such miraculous fine tuning could also be a hint that some deeper physical mechanism is at work. That is, either some previously unknown symmetry is protecting the Higgs mass or the cutoff scale is $\mathcal{O}\left(\SI{}{\tera\electronvolt}\right)$\fxnote{explain better and cite something}.

@@

\section{Long-Lived Particles}
    The SUSY variants most relevant to this discussion are those that are motivated by naturalness and \fxnote{and/or?} produce a signal that could be missed by common SUSY search techniques. Specifically, we are interested in SUSY models that predict new long-lived particles (LLPs), which are defined as particles whose lifetimes are such that they decay at least a detectable distance from the nominal interaction point. This category includes everything from particles that decay at 100um to particles that propagate through the entire detector\fxnote{I made up 100um. Find real number and cite source}.

    In the SM and in BSM theories, LLPs result from any of the following conditions: (1) a symmetry that guarantees stability, (2) weak couplings to decay products, (3) highly-virtual intermediate states, or (4) limited phase space in which to decay\fxnote{find source to cite}. In the SM, charge and baryon number conservation guarantee the absolute stability of the electron and proton as the the lightest particles carrying their respective conserved quantities. Neutrons and muons, whose decays are shown in Fig.\ref{sm_llp_processes}, are long-lived for reasons (2), (3), and (4). Both are pure weak decays, which gives them a smaller coupling constant, but they also decay through a W-boson, which is far more massive than either parent particle, and most importantly, the proton and electron masses leave them little phase space in which to decay\fxnote{find something to cite}. \fxnote{Mention neutrino, pion, kaon, some other bound state I might be forgetting?}

@@

    On the other hand, models with weakly coupled R-parity violating (RPV) terms produce long-lived but not perfectly stable LSPs\fxnote{find something to cite}. A similar situation arises in GMSB models\fxnote{should I describe GMSB?} where the gravitino is the LSP. The coupling between the next-to-LSP (NLSP) and gravitino is inversely proportional to the scale of SUSY breaking, which supresses the NSLP decay rate and can lead to LLPs~\cite{dimopoulos_low_energy}.

@@

\section{The CMS Detector}\fxnote{Add pictures of everything}
    \subsection{Overview}
        \fxnote{define eta, coordinate system in general}
        \fxnote{Mention that ATLAS is similar, says Andrew}  
        \fxnote{Maybe mention PDFs. Definitely know about them, says Andrew}  
        The Compact Muon Solenoid (CMS) detector is designed to reconstruct a variety of particles from collisions at the Large Hadron Collider (LHC), which is a circular proton-proton collider that ran with center-of-mass energy \SI{7}{\tera\electronvolt} in 2010-11, \SI{8}{\tera\electronvolt} in 2012, and \SI{13}{\tera\electronvolt} from 2015 to the present\fxnote{say better}\fxnote{cite something?}. To meet this goal, CMS combines information from several subdetectors nested radially in and around a \SI{4}{\tesla} superconducting solenoid. All together, CMS is \SI{21.6}{\m} long, \SI{14.6}{\m} across, and weighs approximately \SI{12500}{t}. The subdetectors that make up CMS are shown in Fig.~\ref{cms} and described below~\cite{cms_experiment}.

        \noindent \begin{figure}[htbp] \begin{center}
        \includegraphics[width=0.7\textwidth]{images/cms}
            \caption{The CMS detector\fxnote{Cite that website from Andrew's paper}}
        \label{cms}
        \end{center} \end{figure}

    \subsection{Tracker}
        \fxnote{cite CMS TDR, jinst 12 c02033}
        \fxnote{mention that analyses below used pre-upgrade tracker}
        In the region closest to the interaction point, CMS uses a fast, high-granularity silicon tracker to reconstruct tracks and pinpoint primary and secondary vertices. The tracker is subdivided into two regions. Inside \SI{20}{cm}, the pixel tracker  uses smaller\fxnote{than what?} silicon pixels to keep the occupancy near \SI{0.1}{\percent} despite the high\fxnote{specify} particle flux. Outside \SI{20}{\cm}, the silicon strip tracker (SST) achieves similar occupancy with larger silicon strips~\cite{cms_tdr}.\fxnote{check occupacy numbers}

@@

        \noindent \begin{figure}[htbp] \begin{center}
            \includegraphics[width=0.4\textwidth]{images/pixel_old_new}
            \caption{Diagram showing barrel pixel layers before and after the winter 2016/17 upgrade.\fxnote{cite website found through Andrew's citations}}
            \label{pixel_old_new}
        \end{center} \end{figure}

@@

    \subsection{Electromagnetic Calorimeter}
        \fxnote{cite CMS TDR}
        \fxnote{I define EE \& EB but never use those definitions}
        After traversing the inner tracker, particles next encounter the electromagnetic calorimeter (ECAL). As a homogeneous scintillation calorimeter, ECAL uses \num{61200} lead tunstate crystals in the barrel and \num{7324} in each endcap to reconstruct the energy deposited during electromagnetic showers. Lead tungstate crystals allow for a fast (\SI{80}{\percent} of light within \SI{25}{ns}), compact (radiation length = \SI{0.89}{cm}), fine-grained (Moli\`ere radius = \SI{2.2}{cm}), and radiation hard (up to 10 Mrad\fxnote{learn context for this number}) calorimeter. The main drawback is the relatively low light yield (\SI{30}{photon\per\mega\electronvolt}), which necesitates photodetectors with intrinsic gain that work in magnetic fields\cite{cms_experiment, cms_tdr}.

@@

    \subsection{Hadronic Calorimeter}
        \fxnote{cite CMS TDR}
        \fxnote{define HB \& HF never to be used}
        The hadronic calorimeter (HCAL) uses \SI{3.7}{mm} plates of plastic scintillator interspersed within brass absorber to reconstruct the energy deposited during hadronic showers. Embedded wavelength-shifting fibers capture the scintillation light and transfer it to clear fibers to be read out by hybrid photodiodes\cite{cms_experiment, cms_tdr}.

@@

    \subsection{Solenoid}
        \fxnote{cite cms experiment at the cern lhc}
        CMS employs a \SI{4}{T} superconducting solenoid to aid in the measurement of charged particle momenta. The solenoid consists of \num{2168} turns of Nb-Ti superconducting cable and has an inner diameter of \SI{6}{m} and a length of \SI{12.5}{m}. The flux returns through an iron yoke that also houses the muon system. The significant \fxnote{quantify?} bending power helps CMS meet its physics goals of unambiguous sign determination and momentum resolution of \SI{10}{\percent} for muons up to \SI{1}{\tera\electronvolt}\cite{cms_experiment}.

    \subsection{Muon System}
        \fxnote{cite CMS TDR}
        \fxnote{Specifically mention CMS design specs to meet physics goals?}
        The muon system (MS), which is the outermost subdetector, detects muons with \num{3} varieties of gaseous ionization detectors: drift tubes (DTs) in the barrel, cathode strip chambers (CSCs) in the endcaps, and resistive plate chambers (RPCs) in both the barrel and endcaps. The choice of detector varies between barrel and endcap due to the difference in neutron-induced background\fxnote{what?} , muon rates, and magnetic field strength\fxnote{why does B matter?}. The RPCs are useful in the endcaps and barrel because their fast response and good time resolution allow unambiguous bunch crossing identification. Combining MS and tracker measurements improves the muon pT\fxnote{define} resolution at high momenta where the tracker $\pT$ resolution declines \fxnote{quantify or show std plot?}\cite{cms_experiment, cms_tdr}.

@@

    \subsection{Trigger}
        \fxnote{cite CMS TDR, jinst 12 c03021}
        \fxnote{Is entire detector read out at L1?}
        The trigger reduces the data writing rate from the \SI{40}{\mega\hertz} collision rate to less than \SI{1}{\kilo\hertz} so that events can be written to tape. The rate reduction happens in two stages: Level-1 (L1) and High-Level Trigger (HLT). L1 uses input from ECAL, HCAL, and MS that represents physics objects ($\mathrm{e}$/$\gamma$, $\tau$, jets, $\mu$, and $\MET$) and uses custom electronics to reduce the rate to $\approx$100 kHz in 3.8 us. The HLT then uses a dedicated processor farm to further reduce the rate to the desired $<$ 1 kHz\cite{cms_experiment, cms_trigger_upgrade}.

\section{LHC Searches for Long-lived Particles} \label {analyses}
    \fxnote{mention which data was used for each search} 
    \subsection{Overview}
        LLPs leave distinct signatures that differentiate them from SM backgrounds but also add complexity to LLP analyses. CMS and ATLAS were mainly designed to reconstruct stable particles from prompt decays\fxnote{find something to cite}, so LLP searches are often forced\fxnote{the searchers are forced, not the searches} to stray from the traditional analysis path in terms of object reconstruction\fxnote{really reconstruction? Maybe object definition?} and background estimation. This report focuses on four complementary LLP signatures: displaced vertices, disappearing tracks, low-velocity particles, and particles that come to a stop before decaying. See Fig.~\ref{jamie_cartoon} for a cartoon of several possible LLP signatures\fxnote{probably a better way to word this sentence}.

@@

    \subsection{Displaced Vertices}
        \fxnote{cite CONF-2017-026}
        LLPs that decay before reaching the inner tracker can be identified with a displaced vertex (DV)\fxnote{DVs defined clearly enough in overview?}\fxnote{Is it still a DV if it decays in active material somewhere?}. ATLAS performs a search for such particles by looking for at least one DV along with large MET and large track multiplicity~\cite{atlas_displaced}.\fxnote{make sure I can explain why met and multiplicity} The search is sensitive to LLPs with lifetimes in the \SI{20}{\pico\s} to \SI{10}{\nano\s} range\fxnote{Am I defining sensitivity range consistently across analyses?}.

        This search\fxnote{repetative wording} uses a simplified benchmark model in the Split SUSY paradigm where gluinos are kinematically accessible but squarks are not. The relevant process, shown in Fig.~\ref{displaced_process}, is pair-produced gluinos that decay to a quark-\fxnote{how to punctuate?}virtual squark pair. The virtual squark then decays to a quark and the LSP, presumed to be a neutralino\fxnote{have I defined neutralino?}\fxnote{why is this a reasonable assumption?} \fxnote{steal diagram from andrew's preapproval. I can't access it - ask andrew}. The large mass of the intermediate squark suppresses the gluino decay, giving the gluino enough time to form an R-hadron\fxnote{first use. Define here if not in LLP section} that propagates some measureable distance before decaying.

@@

        \noindent \begin{figure}[htbp] \begin{center}
        \begin{subfigure}[htbp]{0.45\textwidth} \begin{center}
        \includegraphics[width=\textwidth]{images/displaced_cross_section_limit}
        \end{center} \end{subfigure}
        \begin{subfigure}[htbp]{0.45\textwidth} \begin{center}
        \includegraphics[width=\textwidth]{images/displaced_mass_limit}
        \end{center} \end{subfigure}
        \caption{Limit\fxnote{explain}.~\cite{atlas_displaced}}
        \label{displaced_limit}
        \end{center} \end{figure}

    \subsection{Disappearing Tracks}
        \fxnote{cite CONF-2017-017}
        Certain models with nearly mass-degenerate LSP and NLSP have the potential to present a remarkable signal: charged particles that seem to disappear in the tracker. Specifically, models with winos as the lightest gauginos\fxnote{have these words been defined?}, such as many AMSB models\fxnote{why amsb? what is amsb?}, will have a chargino NLSP with an observable lifetime (\SI{0.2}{\nano\s} is a typical estimate\fxnote{why? how typical?}) that eventaully decays to the neutralino LSP and, commonly, a pion\fxnote{why?}. The nearly-degenerate masses have two important phenomenological roles: (1) increase the chargino lifetime, and (2) ensure low-momentum decay products. The low-momentum pion is then unlikely to be reconstructed, which results in a disappearing track~\cite{atlas_disappearing}.

@@

        \fxnote{ugh, this is too hard. just steal diagram from paper or make in jaxodraw. Maybe even use simulated event instead, but make sure I can draw these if asked?}  
        \noindent \begin{figure}[htbp] \begin{center}
        \begin{subfigure}[htbp]{0.2\textwidth} \begin{center}
        \includegraphics[width=\textwidth]{images/disappearing_ew_process}
        \end{center} \end{subfigure}
        \qquad
        \begin{subfigure}[htbp]{0.2\textwidth} \begin{center}
        \includegraphics[width=\textwidth]{images/disappearing_qcd_process}
        \end{center} \end{subfigure}
        \fxnote{Can't get it to work in tikz-feynman. Just take from paper for now.} 
        %\feynmandiagram [layered layout] {
        %    i1 -- [plain] a [blob],
        %    i2 -- [plain] a,
        %    a -- [plain, boson] b [blob],
        %    a -- [plain] f1,
        %    a -- [plain, boson] f2,
        %    b -- [plain, boson] f3,
        %    b -- [scalar] f4,
        %};
            \caption{Processes\fxnote{explain}.~\cite{atlas_disappearing}}
        \label{disappearing_processes}
        \end{center} \end{figure}

@@

        No excess over expected background is observed, so upper limits are placed in the NLSP chargino mass-lifetime plane for the EW signal process and in the NLSP chargino mass-gluon mass plane for the QCD production process \fxnote{hopefully will be way clearer after texing in sparticle symbols}. Fig.~\ref{disappearing_limits} shows the limits for both processes.

        \noindent \begin{figure}[htbp] \begin{center}
        \begin{subfigure}[htbp]{0.3\textwidth} \begin{center}
        \includegraphics[width=\textwidth]{images/disappearing_ew_limit}
        \end{center} \end{subfigure}
        \begin{subfigure}[htbp]{0.3\textwidth} \begin{center}
        \includegraphics[width=\textwidth]{images/disappearing_qcd_200ps_limit}
        \end{center} \end{subfigure}
        \begin{subfigure}[htbp]{0.3\textwidth} \begin{center}
        \includegraphics[width=\textwidth]{images/disappearing_qcd_1ns_limit}
        \end{center} \end{subfigure}
            \caption{Limits\fxnote{explain}.~\cite{atlas_disappearing}}
        \label{disappearing_limits}
        \end{center} \end{figure}

@@

    \subsection{Heavy Stable Charged Particles}
        \fxnote{nice dE/dx plot (fig 1)}
        Particles that live for more than a couple \SI{}{ns}\fxnote{Check this and be more exact} will appear stable in CMS and ATLAS because they can traverse the entire detector before decaying. If such a particle is also sufficiently massive, it can be catigorized as a heavy stable charged particle (HSCP). HSCPs are characterterized by a greater rate of energy loss via ionization due to the lower speed with which they pass through detector material\fxnote{redundant w/ overview. Probably better explained here than there}. Additionally, HSCPs can be fractionally, singly, or multiply charged, which also affects their rate of energy loss, as shown in Fig.~\ref{hscp_dedx}~\cite{cms_hscp}.

        \noindent \begin{figure}[htbp] \begin{center}
        \includegraphics[width=0.5\textwidth]{images/hscp_dedx}
            \caption{Ionization energy loss as a function of momentum for paticles of varying mass and charge~\cite{cms_hscp}}
        \label{hscp_dedx}
        \end{center} \end{figure}

@@

        \fxnote{I show seperate trk, trk+tof limits but never explain that in text} 
        \noindent \begin{figure}[htbp] \begin{center}
        \begin{subfigure}[htbp]{0.45\textwidth} \begin{center}
        \includegraphics[width=\textwidth]{images/hscp_limit_tracker_only}
        \end{center} \end{subfigure}
        \begin{subfigure}[htbp]{0.45\textwidth} \begin{center}
        \includegraphics[width=\textwidth]{images/hscp_limit_tracker_tof}
        \end{center} \end{subfigure}
        \caption{Production cross section limits on several particles as a function of mass for the tracker-only and tracker+TOF analyses~\cite{cms_hscp}}
        \label{hscp_limits}
        \end{center} \end{figure}

@@

    Determining the acceptable degree of fine tuning is even more open ended. Before null results and serious constraints started rolling in, it was common\fxnote{feng} to refer to the $\Delta \lesssim 10$ regions of parameter space as natural, but as models are constrained, physicists seem willing to accept greater degrees of fine tuning. Feng proposes an interesting way to recast this problem: instead of asking which regions of parameter space are most natural, physicists should instead ask themselves, ``If superpartners were discovered, what level of fine fine-tuning would be sufficient to convince me that the gauge hierarchy problem is solved by sypersymmetry and I should move on to researhing other problems?"\fxnote{paraphrase for brevity?} Anecdotally, acceptable levels of fine tuning tend to increase when the question is posed in this way~\cite{feng}.

    Exact upper limits on natural sparticle masses are therefore hard to come by. A reasonable and somewhat typical set of values is given in~\cite{drees_kim}: $m_{\tilde{H}} \lesssim \SI{500}{\giga\electronvolt}$, $m_{\tilde{t}} \lesssim \SI{1.5}{\tera\electronvolt}$, and $m_{\tilde{g}} \lesssim \SI{3}{\tera\electronvolt}$. Fig.~\ref{atlas_susy_summary} summarizes current SUSY limits from ATLAS searches (CMS limits are generally similar, but a recent plot was not available). Upon inspection, one sees that experimental limits are starting to push on naturalness bounds in many scenarios.

>>>>

    SM parameters, such as masses and couplings, are corrected by interactions with virtual particles through loop diagrams. Fig.~\ref{loop_diagrams}, for example, shows interactions that correct the Higgs and electron masses. Because the Higgs is a fundamental scalar, its corrections depend quadratically on the cutoff scale, $\Lambda$, which is the largest energy scale for which the SM is valid. The fermions and gauge bosons, on the other hand, only depend logarithmically\fxnote{be able to work out example} on $\Lambda$~\cite{dine_naturalness}. The next section explores the concept of naturalness, a major motivating principle for beyond the standard model (BSM) physics that is directly tied to the relationship between the bare Higgs mass and the terms that correct it.

@@

    All massive particles correct the Higgs mass with terms proportional to $\Lambda^2$~\cite{dine_naturalness}\fxnote{show diagrams and terms?}. Experimentally, the Higgs mass is \SI{125}{\giga\electronvolt}~\cite{cms_higgs, atlas_higgs}, a value that must be achieved by the sum of the bare mass and all corrections. Although corrections differ in sign, the precise cancellation of many extremely large terms to produce a value as small as the Higgs mass is extremely unlikely. In fact, the fine-tuning necessary to produce the Higgs mass is on the order of one part in \num{e30} if the SM is valid up to the Plank scale ($\mPlanck\approx$ \SI{e16}{\giga\electronvolt}), where quantum gravity becomes important~\cite{giudice_naturally}. While it's possible that the bare mass and corrections just so happen to almost perfectly cancel, such miraculous fine tuning could also be a hint that some deeper physical mechanism is at work.

@@

\section{Long-Lived Particles}
    The SUSY variants most relevant to this discussion are those that are motivated by naturalness and \fxnote{and/or?} produce a signal that could be missed by common SUSY search techniques. Specifically, we are interested in SUSY models that predict new long-lived particles (LLPs), which are defined as particles whose lifetimes are such that they decay at least a detectable distance from the nominal interaction point. This category includes everything from particles that decay before reaching the inner tracker to particles that propagate through the entire detector.

    In the SM and in BSM theories, LLPs result from any of the following conditions: (1) a symmetry that guarantees stability, (2) weak couplings to decay products, (3) highly-virtual intermediate states, or (4) limited phase space in which to decay. In the SM, charge and baryon number conservation guarantee the absolute stability of the electron and proton as the the lightest particles carrying their respective conserved quantities. Neutrons and muons, whose decays are shown in Fig.\ref{sm_llp_processes}, are long-lived for reasons (2), (3), and (4). Both are pure weak decays, which gives them a smaller coupling constant, but they also decay through a W-boson, which is far more massive than either parent particle, and most importantly, the proton and electron masses leave them little phase space in which to decay. \fxnote{Mention neutrino, pion, kaon, some other bound state I might be forgetting?}

@@

    On the other hand, models with weakly coupled R-parity violating (RPV) terms produce long-lived but not perfectly stable LSPs~\cite{fate}. A similar situation arises in GMSB models\fxnote{should I describe GMSB?} where the gravitino is the LSP. The coupling between the next-to-LSP (NLSP) and gravitino is inversely proportional to the scale of SUSY breaking, which supresses the NSLP decay rate and can lead to LLPs~\cite{dimopoulos_low_energy}.

@@

\section{The CMS Detector}\fxnote{Add pictures of everything}
    \subsection{Overview}
        \fxnote{define eta, coordinate system in general}
        \fxnote{Mention that ATLAS is similar, says Andrew}  
        The Compact Muon Solenoid (CMS) detector is designed to reconstruct a variety of particles from collisions at the Large Hadron Collider (LHC), which is a circular proton-proton collider that ran with center-of-mass energy \SI{7}{\tera\electronvolt} in 2010-11, \SI{8}{\tera\electronvolt} in 2012, and \SI{13}{\tera\electronvolt} from 2015 to the present\fxnote{say better}. To meet this goal, CMS combines information from several subdetectors nested radially in and around a \SI{4}{\tesla} superconducting solenoid. All together, CMS is \SI{21.6}{\m} long, \SI{14.6}{\m} across, and weighs approximately \SI{12500}{t}. The subdetectors that make up CMS are shown in Fig.~\ref{cms} and described below~\cite{cms_experiment}.

        \noindent \begin{figure}[htbp] \begin{center}
        \includegraphics[width=0.7\textwidth]{images/cms}
        \caption{The CMS detector~\cite{cms_image}}
        \label{cms}
        \end{center} \end{figure}

    \subsection{Tracker}
        \fxnote{mention that analyses below used pre-upgrade tracker}
        In the region closest to the interaction point, CMS uses a fast, high-granularity silicon tracker to reconstruct tracks and pinpoint primary and secondary vertices. The tracker is subdivided into two regions. Inside \SI{20}{cm}, the pixel tracker  uses smaller\fxnote{than what?} silicon pixels to keep the occupancy near \SI{0.1}{\percent} despite the high\fxnote{specify} particle flux. Outside \SI{20}{\cm}, the silicon strip tracker (SST) achieves similar occupancy with larger silicon strips~\cite{cms_tdr}.\fxnote{check occupacy numbers}

@@

        \noindent \begin{figure}[htbp] \begin{center}
            \includegraphics[width=0.4\textwidth]{images/pixel_old_new}
            \caption{Diagram showing barrel pixel layers before and after the winter 2016/17 upgrade.~\cite{pixel_old_new}}
            \label{pixel_old_new}
        \end{center} \end{figure}

@@

    \subsection{Electromagnetic Calorimeter}
        \fxnote{I define EE \& EB but never use those definitions}
        After traversing the inner tracker, particles next encounter the electromagnetic calorimeter (ECAL). As a homogeneous scintillation calorimeter, ECAL uses \num{61200} lead tunstate crystals in the barrel and \num{7324} in each endcap to reconstruct the energy deposited during electromagnetic showers. Lead tungstate crystals allow for a fast (\SI{80}{\percent} of light within \SI{25}{ns}), compact (radiation length = \SI{0.89}{cm}), fine-grained (Moli\`ere radius = \SI{2.2}{cm}), and radiation hard (up to 10 Mrad\fxnote{learn context for this number}) calorimeter. The main drawback is the relatively low light yield (\SI{30}{photon\per\mega\electronvolt}), which necesitates photodetectors with intrinsic gain that work in magnetic fields\cite{cms_experiment, cms_tdr}.

@@

    \subsection{Hadronic Calorimeter}
        \fxnote{define HB \& HF never to be used}
        The hadronic calorimeter (HCAL) uses \SI{3.7}{mm} plates of plastic scintillator interspersed within brass absorber to reconstruct the energy deposited during hadronic showers. Embedded wavelength-shifting fibers capture the scintillation light and transfer it to clear fibers to be read out by hybrid photodiodes\cite{cms_experiment, cms_tdr}.

@@

    \subsection{Solenoid}
        CMS employs a \SI{4}{T} superconducting solenoid to aid in the measurement of charged particle momenta. The solenoid consists of \num{2168} turns of Nb-Ti superconducting cable and has an inner diameter of \SI{6}{m} and a length of \SI{12.5}{m}. The flux returns through an iron yoke that also houses the muon system. The significant \fxnote{quantify?} bending power helps CMS meet its physics goals of unambiguous sign determination and momentum resolution of \SI{10}{\percent} for muons up to \SI{1}{\tera\electronvolt}\cite{cms_experiment}.

    \subsection{Muon System}
        \fxnote{Specifically mention CMS design specs to meet physics goals?}
        The muon system (MS), which is the outermost subdetector, detects muons with \num{3} varieties of gaseous ionization detectors: drift tubes (DTs) in the barrel, cathode strip chambers (CSCs) in the endcaps, and resistive plate chambers (RPCs) in both the barrel and endcaps. The choice of detector varies between barrel and endcap due to the difference in neutron-induced background\fxnote{what?} , muon rates, and magnetic field strength\fxnote{why does B matter?}. The RPCs are useful in the endcaps and barrel because their fast response and good time resolution allow unambiguous bunch crossing identification. Combining MS and tracker measurements improves the muon pT\fxnote{define} resolution at high momenta where the tracker $\pT$ resolution declines \fxnote{quantify or show std plot?}\cite{cms_experiment, cms_tdr}.

@@

    \subsection{Trigger}
        \fxnote{Is entire detector read out at L1?}
        The trigger reduces the data writing rate from the \SI{40}{\mega\hertz} collision rate to less than \SI{1}{\kilo\hertz} so that events can be written to tape. The rate reduction happens in two stages: Level-1 (L1) and High-Level Trigger (HLT). L1 uses input from ECAL, HCAL, and MS that represents physics objects ($\mathrm{e}$/$\gamma$, $\tau$, jets, $\mu$, and $\MET$) and uses custom electronics to reduce the rate to $\approx$100 kHz in 3.8 us. The HLT then uses a dedicated processor farm to further reduce the rate to the desired $<$ 1 kHz\cite{cms_experiment, cms_trigger_upgrade}.

\section{LHC Searches for Long-lived Particles} \label {analyses}
    \fxnote{mention which data was used for each search} 
    \subsection{Overview}
        LLPs leave distinct signatures that differentiate them from SM backgrounds but also add complexity to LLP analyses. CMS and ATLAS were mainly designed to reconstruct stable particles from prompt decays, so LLP searches are often forced\fxnote{the searchers are forced, not the searches} to stray from the traditional analysis path in terms of object reconstruction and background estimation. This report focuses on four complementary LLP signatures: displaced vertices, disappearing tracks, low-velocity particles, and particles that come to a stop before decaying. See Fig.~\ref{jamie_cartoon} for a cartoon of several possible LLP signatures\fxnote{probably a better way to word this sentence}.

@@

    \subsection{Displaced Vertices}
        LLPs that decay before reaching the inner tracker can be identified with a displaced vertex (DV)\fxnote{DVs defined clearly enough in overview?}\fxnote{Is it still a DV if it decays in active material somewhere?}. ATLAS performs a search for such particles by looking for at least one DV along with large MET and large track multiplicity~\cite{atlas_displaced}.\fxnote{make sure I can explain why met and multiplicity} The search is sensitive to LLPs with lifetimes in the \SI{20}{\pico\s} to \SI{10}{\nano\s} range\fxnote{Am I defining sensitivity range consistently across analyses?}.

        This search\fxnote{repetative wording} uses a simplified benchmark model in the Split SUSY paradigm where gluinos are kinematically accessible but squarks are not. The relevant process, shown in Fig.~\ref{displaced_process}, is pair-produced gluinos that decay to a quark-\fxnote{how to punctuate?}virtual squark pair. The virtual squark then decays to a quark and the LSP, presumed to be a neutralino\fxnote{have I defined neutralino?}\fxnote{why is this a reasonable assumption?}. The large mass of the intermediate squark suppresses the gluino decay, giving the gluino enough time to form an R-hadron\fxnote{first use. Define here if not in LLP section} that propagates some measureable distance before decaying.

@@

        \noindent \begin{figure}[htbp] \begin{center}
        \begin{subfigure}[htbp]{0.45\textwidth} \begin{center}
        \includegraphics[width=\textwidth]{images/displaced_cross_section_limit}
        \end{center} \end{subfigure}
        \begin{subfigure}[htbp]{0.45\textwidth} \begin{center}
        \includegraphics[width=\textwidth]{images/displaced_mass_limit}
        \end{center} \end{subfigure}
            \caption{Limits on the gluino pair-production cross section (left) and gluino mass as a function of lifetime (right)~\cite{atlas_displaced}.}
        \label{displaced_limit}
        \end{center} \end{figure}

    \subsection{Disappearing Tracks}
        Certain models with nearly mass-degenerate LSP and NLSP have the potential to present a remarkable signal: charged particles that seem to disappear in the tracker. Specifically, models with winos as the lightest gauginos\fxnote{have these words been defined?}, such as many AMSB models\fxnote{why amsb? what is amsb?}, will have a chargino NLSP with an observable lifetime (\SI{0.2}{\nano\s} is a typical estimate\fxnote{why? how typical?}) that eventaully decays to the neutralino LSP and, commonly, a pion\fxnote{why?}. The nearly-degenerate masses have two important phenomenological roles: (1) increase the chargino lifetime, and (2) ensure low-momentum decay products. The low-momentum pion is then unlikely to be reconstructed, which results in a disappearing track~\cite{atlas_disappearing}.

@@

        \fxnote{ugh, this is too hard. just steal diagram from paper or make in jaxodraw. Maybe even use simulated event instead, but make sure I can draw these if asked?}  
        \noindent \begin{figure}[htbp] \begin{center}
        \begin{subfigure}[htbp]{0.2\textwidth} \begin{center}
        \includegraphics[width=\textwidth]{images/disappearing_ew_process}
        \end{center} \end{subfigure}
        \qquad
        \begin{subfigure}[htbp]{0.2\textwidth} \begin{center}
        \includegraphics[width=\textwidth]{images/disappearing_qcd_process}
        \end{center} \end{subfigure}
        \fxnote{Can't get it to work in tikz-feynman. Just take from paper for now.} 
        %\feynmandiagram [layered layout] {
        %    i1 -- [plain] a [blob],
        %    i2 -- [plain] a,
        %    a -- [plain, boson] b [blob],
        %    a -- [plain] f1,
        %    a -- [plain, boson] f2,
        %    b -- [plain, boson] f3,
        %    b -- [scalar] f4,
        %};
            \caption{Benchmark signal processes: electroweak gaugino production (left) and gluino pair-production (right)~\cite{atlas_disappearing}.}
        \label{disappearing_processes}
        \end{center} \end{figure}

@@

        No excess over expected background is observed, so upper limits are placed in the NLSP chargino mass-lifetime plane for the EW signal process and in the NLSP chargino mass-gluon mass plane for the QCD production process. Fig.~\ref{disappearing_limits} shows the limits for both processes.

        \noindent \begin{figure}[htbp] \begin{center}
        \begin{subfigure}[htbp]{0.3\textwidth} \begin{center}
        \includegraphics[width=\textwidth]{images/disappearing_ew_limit}
        \end{center} \end{subfigure}
        \begin{subfigure}[htbp]{0.3\textwidth} \begin{center}
        \includegraphics[width=\textwidth]{images/disappearing_qcd_200ps_limit}
        \end{center} \end{subfigure}
        \begin{subfigure}[htbp]{0.3\textwidth} \begin{center}
        \includegraphics[width=\textwidth]{images/disappearing_qcd_1ns_limit}
        \end{center} \end{subfigure}
            \caption{Limits on chargino mass and lifetime in the electroweak production channel (left) and gluino and chargino masses in the gluon pair-production channel with a \SI{0.2}{\nano\s} (center) and \SI{1.0}{\nano\s} chargino lifetime.~\cite{atlas_disappearing}}
        \label{disappearing_limits}
        \end{center} \end{figure}

@@

    \subsection{Heavy Stable Charged Particles}
        Particles that live for more than a couple \SI{}{ns}\fxnote{Check this and be more exact} will appear stable in CMS and ATLAS because they can traverse the entire detector before decaying. If such a particle is also sufficiently massive, it can be catigorized as a heavy stable charged particle (HSCP). HSCPs are characterterized by a greater rate of energy loss via ionization due to the lower speed with which they pass through detector material\fxnote{redundant w/ overview. Probably better explained here than there}. Additionally, HSCPs can be fractionally, singly, or multiply charged, which also affects their rate of energy loss, as shown in Fig.~\ref{hscp_dedx}~\cite{cms_hscp}.

        \noindent \begin{figure}[htbp] \begin{center}
        \includegraphics[width=0.4\textwidth]{images/hscp_dedx}
            \caption{Ionization energy loss as a function of momentum for paticles of varying mass and charge~\cite{cms_hscp}}
        \label{hscp_dedx}
        \end{center} \end{figure}

@@

        \fxnote{I show seperate trk, trk+tof limits but never explain that in text} 
        \noindent \begin{figure}[htbp] \begin{center}
        \begin{subfigure}[htbp]{0.4\textwidth} \begin{center}
        \includegraphics[width=\textwidth]{images/hscp_limit_tracker_only}
        \end{center} \end{subfigure}
        \begin{subfigure}[htbp]{0.4\textwidth} \begin{center}
        \includegraphics[width=\textwidth]{images/hscp_limit_tracker_tof}
        \end{center} \end{subfigure}
        \caption{Production cross section limits on several particles as a function of mass for the tracker-only and tracker+TOF analyses~\cite{cms_hscp}}
        \label{hscp_limits}
        \end{center} \end{figure}

@@

    Determining the acceptable degree of fine tuning is even more subjective. Before null results and serious constraints started rolling in, it was common to refer to the $\Delta \lesssim 10$ regions of parameter space as natural, but as models are constrained, physicists seem willing to accept greater degrees of fine tuning. Feng proposes an interesting way to recast this problem: instead of asking which regions of parameter space are most natural, physicists should instead ask themselves, ``If superpartners were discovered, what level of fine fine-tuning would be sufficient to convince me that the gauge hierarchy problem is solved by sypersymmetry and I should move on to researching other problems?"\fxnote{paraphrase for brevity?} Anecdotally, fine tuning thresholds tend to increase when the question is posed in this way~\cite{feng}.

    Exact upper limits on natural sparticle masses are therefore hard to come by. A somewhat typical set of values is given in~\cite{drees_kim}: $m_{\tilde{H}} \lesssim \SI{500}{\giga\electronvolt}$, $m_{\tilde{t}} \lesssim \SI{1.5}{\tera\electronvolt}$, and $m_{\tilde{g}} \lesssim \SI{3}{\tera\electronvolt}$. Fig.~\ref{atlas_susy_summary} summarizes current SUSY limits from ATLAS searches (CMS limits are generally similar, but a recent plot is not available). Upon inspection, one sees that experimental limits are starting to push on naturalness bounds in many scenarios.

    \fxnote{what do I cite?}  
    \noindent \begin{figure}[htbp] \begin{center}
    \includegraphics[width=0.7\textwidth]{images/atlas_susy_summary}
    \caption{Summary of current experimental limits on SUSY models from ATLAS searches.}
    \label{atlas_susy_summary}
    \end{center} \end{figure}
